Stabilizing selection, archaic introgression, and the distribution of shared functional variation

### 2023-07-05 doc/manuscript.tex
update some figures

Genomic studies have shown that admixture is a common occurrance between
diverged populations and closely related taxa \citep{}, and potentially has
been widespread in primate \citep{} and hominin \citep{} evolution.
Understanding the effects of admixture on phenotypic and molecular variation is
therefore relevant to many natural systems, including to studies of the genetic
basis of complex traits and diseases. In particular, archaic introgression in
\emph{Homo sapiens} involving Neanderthals and Denisovans has recently
attracted considerable attention, both in understanding the historical
processes leading to observed distributions of introgressed DNA in present-day
populations \citep{prufer2014complete,villanea20??,chen2020identifying} as well
as its contribution to quantitative traits \citep{sankararaman,wei,others}.

After their introduction through admixture, introgressed alleles may be
selected either for or against. Some introgressed haplotypes appear to have
been positively selected in modern humans
\citep{huerta2014altitude,gower2021detecting,Enard}, possibly because such
alleles are locally adaptive and provide a fitness advantage when encountering
a novel environment. Despite these cases of adaptive introgression,
introgressed alleles were more likely to have been negatively selected in
modern humans \citep{harris2016genetic,juric2016strength}. As Neanderthal and
Denisovan population sizes are thought to have been relatively small for
hundreds of thousands of years, population genetics theory predicts that they
would accumulate deleterious variation at a faster rate than larger
populations.  Introgressed haplotypes would then carry more deleterious
variants, which would quickly be selected against after admixture. In mapping
the distribution of haplotypes identified as introgressed from Neanderthals,
there is a reduction of Neanderthal-related ancestry in enhancers and
regulatory regions
\citep{petr2019limits,telis2020selection,yermakovich2023long}. This
Neanderthal-allelic depletion, or ``deserts,'' support the hypothesis that
introgressed functional alleles were selected against
\citep{sankararaman2014genomic,sankararaman2016combined}.

There is genetic evidence that early \emph{H. sapiens} reciprocally contributed
to Neanderthal genomes \citep{kuhlwilm2016ancient,hubisz2020mapping}, with
introgression in the reverse direction occurring tens or hundreds of thousands
of years before the Neanderthal-to-human event surrounding the out-of-Africa
expansion around 60 thousand years ago (ka). This is supported by early ``near
modern'' \emph{H.  sapiens} outside of African around 120--100ka or earlier
\citep{schwarcz1988esr,grun2005u,beyer2021climatic}, potentially overlapping
with Neanderthals and providing opportunities for early contacts. While
estimates of the genomic contribution of early \emph{H. sapiens} to
Neanderthals vary \citep{}, up to XX\% of later Neanderthal genomes may have
been contributed through introgression from humans to Neanderthals. Under the
same argument that \emph{H. sapiens}-related haplotypes carry fewer deleterious
alleles, human-introgressed DNA may have been favored in Neanderthal genomes.
The replacement of Neanderthal mitochondrial and Y chromosomes by early human
haplotypes supports this model \citep{posth2017deeply,petr2020evolutionary}.

Models for selection on introgressed alleles are often based on load arguments
(in particular, differences in the rate of accumulation of unconditionally
deleterious variation in populations of different sizes) or hybrid
incompatibilities \citep{mueller,etc}. These arguments are founded primarily in
population genetics theory, eschewing any notion of genetic variation
contributing to phenotypic variation and selection operating on those
phenotypes. However, most phenotypic traits are thought to be under stabilizing
selection \citep{sanjak2018evidence}, including gene regulation
\citep{gilad2006natural,hodgins2015gene,price2022detecting}. This is especially
relevant for selection on Neanderthal-introduced alleles in modern humans, as
the strongest signals of selection against such alleles occur in regulatory
regions \citep{sankararaman2014genomic}.

@@

In two diverged populations, the genomic architecture (i.e., mutations
underlying the trait) of a trait under stabilizing selection has a higher rate
of turnover compared to neutrally evolving loci \citep{yair2022population}. We
might therefore expect an accumulation of fixed differences between the
\emph{H. sapiens} and Neanderthal branches for alleles contributing to a trait,
even when the mean phenotype in each population is close to the same trait
optimum. When alleles that are fixed in one population are introduced to
another population in which they were previously absent, they will be at low
frequency and subsequently selected against. Likewise, if the ancestral allele
is reintroduced to a population fixed for a derived allele, the ancestral
allele will be at low frequency and will be selected against. In either case we
should expect the introgressed allele, whether ancestral or derived, to be
selected against, regardless the historical relative sizes of the populations
involved. This prediction contrasts with the population-genetics ``load''
model, in which haplotypes with fewer deleterious variants (such as those from
the population with larger historical size) are favored after introgression in
either direction.

@@

If all alleles contribute equally to the trait (as \(\pm a\)),
\citet{keightley1988quantitative} showed that the dynamics of $V_G$ can be
approximated with the recursion
\[V_{G,t+1} \approx V_{G,t}\left(1-a^2/2(V_S+V_G)\right)
\left(1-1/2N_e\right) + 2 \mu a^2,\]
where $\mu$ is the per-haploid, per-generation rate of mutation. In the
large-population-size limit, this gives the well-known results
\[\tilde{V_G} \approx 4\mu V_S,\] provided \(V_G \ll V_S\).
Mutations are not expected to each have the same effect size $a$, but
rather will be drawn from some distribution. Here, we will assume mutation
effect sizes are drawn from a normal distribution with mean 0 and given
variance $V_M$.

@@

We model the distribution of allele frequencies underlying the trait using
the approximation that alleles are subject to symmetric underdominant selection.
For an allele with effect size $a$ (in the heterozygous state -- homozygotes
contribute $2a$ to the trait), the selection coefficient
\[s\approx \frac{a^2}{2(V_S + V_G)} \approx \frac{a^2}{2V_S},\]
if $V_G \ll V_S$ \citep[e.g.,][]{simons2018population}. Allelic dynamics can
be modeled using the diffusion approximation, where the expected change
in mean allele frequency per generation is
\[\E[M_{\delta_p}] \approx -s p(1-p)(1-2p),\]
and the expected change in variance of the allele frequency per generation is
\[\E[V_{\delta_p}] \approx \frac{1}{4N_e}p(1-p).\]
From this, because $s$ is always positive for any $a\not=0$, we see that
selection pushes allele frequencies to zero if $p<1/2$ and to one if $p>1/2$.
$p=1/2$ is an unstable equilibrium.

We extend the moments-based solution for the sample site-frequency spectrum
(SFS) \citep{jouganous2017} to include underdominance with given selection
coefficient $s$ (\(=a^2/2(V_S+V_G)\)). The contribution of alleles with effect
size $a$ to the total genetic variance $V_G$ is then found by computing the
expected pairwise diversity from the SFS (with sample size $n$, $\Phi_n$), as
\(V_{G,a} = 2a^2\sum_{j=1}^{n-1}\frac{j(n-j)}{n(n-1)}\Phi_n(j|a,\mu_a)da\).
Here, $\mu_a$ is the mutation rate of alleles with effect size $a$. The total
genetic variance is then \(V_G=\int_{-\infty}^\infty V_{G,a}
\mathcal{N}(0,V_M).\)

@@

Our numerical solution for the SFS allows for non-constant population size
histories, population splits, continuous migration, and admixture. In this
study, we consider relativey simple scenarios involving population splits with
subsequent introgression events. We focus on parameter regimes relevant to
human-Neanderthal history. In the simplest model -- which is not meant to
perfectly match any particular known or inferred history, but instead
demonstrate the effect of reciprocal introgression following population
divergence -- a population of size $N_e=10\,000$ splits into two, on remaining
size $10\,000$ and the other shrinking to size $1\,000$. They remain isolated
for $2N_e$ generations (or 500 thousand years, assuming an average generation
time of 25 years) and then introgression occurs from one branch to the other,
contributing $5\%$ ancestry to the recipient population
(Figure~\ref{toy-admixture}A,B).

@@

In each scenario, we track genetic diversity and phenotypic variance before and
after admixture to understand allele frequency and haplotype dynamics. The
trait optimum is $0$ in all branches, and the strength of selection $V_S=1$
remains constant.

@@

We expect genetic variance to increase after introgression. The amount that
genetic variance increases depends on the allelic differences accumulated
between populations and the effects of those alleles. Assuming no linkage,
dominance or epistasis, \(V_G=\sum_l 2p_l(1-p_l)a_l^2\). After admixture, with
proportion $f$ contributed by the source population (labeled 0) into the focal
population (labeled 1), \(p_l = fp_{l,0} + (1-f)p_{l,1}\). Plugging into the
expression for $V_G$, and after some simple algebra (Appendix~AX), we can
express the expected genetic variance directly after admixture as \[V_G =
fV_{G,0} + (1-f) V_{G,1} + 2f(1-f)\sum_l F_{2,l} a_l^2,\] where $F_2 = (p_0 -
p_1)^2$ is the squared difference in allele frequencies at a locus
\citep[e.g.,][]{peter2016admixture}. This result in known \citep[e.g.,][]{},
showing that additive variance is equal to that in the source populations
weighted by their contributions, plus a term that depends on the divergence at
trait-affecting loci between the populations weighted by the quadratic factor
$2f(1-f)$.

@@

\begin{figure}[t!]
    \centering
    %\includegraphics{}
    \caption{}
    \label{fig:linkage-deviations}
\end{figure}

@@

@@

\begin{itemize}
    \item Stabilizing selection with a constant optimum is an appropriate
        null model for the dynamics of trait variation and selection on
        alleles underlying those traits
    \item Stabilizing selection results in population genetic predictions
        for patterns of introgressed variation in and around functional
        regions that differ from both load- and incompatibility-based arguments
        \begin{itemize}
            \item Load-based models predict that haplotypes from the less ``loaded''
                population will be favored under introgression in either direction
                -- thus, regions of depleted ancestry after introgression in one
                direction will not be expected to also be introgressed ancestry
                deserts after admixture in the opposite direction
            \item \citet{mueller1942} originally pointed out that hybrid
                incompatibilities should not form at single loci -- classical
                incompatibilities require two or more loci, which are not
                expected to be tightly linked to one another. Thus, under a
                model of hybrid incompatibilities, the minor parent ancestry
                is again selected against, but those loci will be in different
                regions. Again, deserts are therefore not expected to overlap
                if selection is largely due to incompatibilities
            \item Instead, stabilizing selection causes introgressed DNA to be
                depleted at the same loci under bi-directional gene flow
                (could be thought of as single-locus incompatibility effects --
                but these are not incompatibilities in the classical sense,
                as the same effects occur even without admixture)
            \item A prediction we could make is that, if current models are correct
                that there has been introgression in both directions between
                \emph{H. sapiens} and Neanderthals, introgressed ancestry deserts
                should overlap with each other more often than expected by chance
        \end{itemize}
    \item Some conjectures about hybrid zones
\end{itemize}

>>>>

Genomic studies have shown that admixture is a common occurrance between
diverged populations and closely related taxa \citep{brandvain2014speciation,
skoglund2015ancient,suvorov2022widespread}, and potentially has been widespread
in primate \citep{tung2017contribution,sorensen2023genome} and hominin
\citep{wolf2018outstanding,peter2020100} evolution. Understanding the effects
of admixture on phenotypic and molecular variation is therefore relevant to
many natural systems, including to studies of the genetic basis of complex
traits and diseases. In particular, archaic introgression in \emph{Homo
sapiens} involving Neanderthals and Denisovans has recently attracted
considerable attention, both in understanding the historical processes leading
to observed distributions of introgressed DNA in present-day populations
\citep{prufer2014complete,villanea2019multiple,chen2020identifying} as well as
its contribution to quantitative traits
\citep{sankararaman2016combined,wei2023lingering}.

After their introduction through admixture, introgressed alleles may be
selected for or against. Some introgressed haplotypes appear to have been
positively selected in modern humans
\citep{huerta2014altitude,racimo2017signatures,enard2018evidence,gower2021detecting},
possibly because such alleles are locally adaptive and provide a fitness
advantage when encountering a novel environment. Despite these cases of
adaptive introgression, introgressed alleles were more likely to have been
negatively selected in modern humans
\citep{harris2016genetic,juric2016strength}. As Neanderthal and Denisovan
population sizes are thought to have been relatively small for hundreds of
thousands of years, theory predicts that they would accumulate deleterious
variation at a faster rate than larger populations. Introgressed haplotypes
would then carry more deleterious variants, which would quickly be selected
against after admixture. \aprcomment{Double check the following statements.} In
mapping the distribution of haplotypes identified as introgressed from
Neanderthals, there is a reduction of Neanderthal-related ancestry in enhancers
and regulatory regions
\citep{petr2019limits,telis2020selection,yermakovich2023long}. This
Neanderthal-allelic depletion, or ``deserts,'' support the hypothesis that
introgressed functional alleles were selected against
\citep{sankararaman2014genomic,sankararaman2016combined}.

There is genetic evidence that early \emph{H. sapiens} reciprocally contributed
to Neanderthal genomes \citep{kuhlwilm2016ancient,hubisz2020mapping}, with
introgression in the reverse direction occurring tens or hundreds of thousands
of years before the Neanderthal-to-human event during the global dispersal of
later humans. This is supported by ``near modern'' \emph{H.  sapiens} outside
of African around 120--100ka or earlier
\citep{schwarcz1988esr,grun2005u,beyer2021climatic}, potentially overlapping
with Neanderthals and providing opportunities for early contacts. While
estimates of the genomic contribution of early \emph{H. sapiens} to
Neanderthals vary \citep{kuhlwilm2016ancient,hubisz2020mapping}, up to 3--5\%
of later Neanderthal genomes may have been contributed through introgression
from humans to Neanderthals. Under the reasoning that \emph{H.
sapiens}-related haplotypes carried fewer deleterious alleles,
human-introgressed DNA would have been favored in Neanderthal genomes. The
replacement of Neanderthal mitochondrial and Y chromosomes by early human
haplotypes supports this model \citep{posth2017deeply,petr2020evolutionary}.

Models for selection against introgressed alleles are often based on load
arguments (in particular, differences in the rate of accumulation of
unconditionally deleterious variation in populations of different sizes) or
hybrid incompatibilities \citep{muller1942isolating}. These arguments are
founded primarily in population genetics theory, with little notion of genetic
variation contributing to phenotypic variation that selection operates on.
Most phenotypic traits are thought to be under stabilizing selection
\citep{sanjak2018evidence,sella2019thinking}, including gene regulation
\citep{gilad2006natural,hodgins2015gene,price2022detecting}. This is relevant
to selection on Neanderthal-introduced alleles in modern humans, as the some of
the strongest signals of selection against such alleles occur in regulatory
regions \citep{sankararaman2014genomic}.

@@

In two diverged populations, the genomic architecture of a trait under
stabilizing selection has a higher rate of turnover compared to neutrally
evolving loci \citep{yair2022population}. We might therefore expect an
accumulation of fixed differences between the \emph{H. sapiens} and Neanderthal
branches for alleles contributing to a trait, even when the mean phenotype in
each population is close to the same trait optimum. When alleles that are fixed
in one population are introduced to another population in which they were
previously absent, they will be at low frequency and subsequently selected
against. Likewise, if the ancestral allele is reintroduced to a population
fixed for a derived allele, the ancestral allele will be at low frequency and
will be selected against. In either case we should expect the introgressed
allele, whether ancestral or derived, to be selected against, regardless the
historical relative sizes of the populations involved. This prediction
contrasts with the population-genetics ``load'' model, in which haplotypes with
fewer deleterious variants (such as those from the population with larger
historical size) are favored after introgression in either direction.

@@

If all alleles contribute equally to the trait (as \(\pm a\)),
\citet{keightley1988quantitative} showed that the dynamics of $V_G$ can be
approximated with the recursion
\[V_{G,t+1} \approx V_{G,t}\left(1-a^2/2(V_S+V_{G,t})\right)
\left(1-1/2N_e\right) + 2 \mu a^2,\]
where $\mu$ is the per-haploid, per-generation rate of mutation. In the
large-population-size limit, this gives the well-known result
\[\tilde{V_G} \approx 4\mu V_S,\] provided \(V_G \ll V_S\).
Mutations are not expected to each have the same effect size $a$, but
rather will be drawn from some distribution. Here, we will assume mutation
effect sizes are drawn from a normal distribution with mean 0 and given
variance $V_M$.

@@

We model the distribution of allele frequencies underlying the trait using
the approximation that alleles are subject to symmetric underdominant
selection \citep{robertson1956effect,keightley1988quantitative}.
For an allele with (heterozygote) additive effect size $a$,
the selection coefficient
\[s\approx \frac{a^2}{2(V_S + V_G)} \approx \frac{a^2}{2V_S},\]
if $V_G \ll V_S$ \citep[e.g.,][]{simons2018population}. Allelic dynamics can
be modeled using the diffusion approximation, where the expected change
in mean allele frequency per generation is
\[\E[M_{\delta_p}] \approx -s p(1-p)(1-2p),\]
and the expected change in variance of the allele frequency per generation is
\[\E[V_{\delta_p}] \approx \frac{1}{4N_e}p(1-p).\]
%From this, because $s$ is always positive for any $a\not=0$, we see that
%selection pushes allele frequencies to zero if $p<1/2$ and to one if $p>1/2$.
%$p=1/2$ is an unstable equilibrium.

We extend the moments-based solution for the sample site-frequency spectrum
(SFS) \citep{jouganous2017} to include underdominance with given selection
coefficient $s$ (\(=a^2/2(V_S+V_G)\)). The contribution of alleles with effect
size $a$ to the total genetic variance $V_G$ is then found by computing the
expected pairwise diversity from the SFS (with sample size $n$, $\Phi_n$), as
\[V_{G,a} = 2a^2\sum_{j=1}^{n-1}\frac{j(n-j)}{n(n-1)}\Phi_n(j|a,\mu_a)da.\]
Here, $\mu_a$ is the mutation rate of alleles with effect size $a$. The total
genetic variance is then 
\[V_G=\int_{-\infty}^\infty V_{G,a}\mathcal{N}(0,V_M).\]

@@

Our numerical solution for the SFS allows for non-constant population size
histories, population splits, continuous migration, and admixture. In this
study, we consider relatively simple scenarios involving population splits with
subsequent introgression events. We focus on parameter regimes relevant to
human-Neanderthal history. In the simplest model -- which is not meant to
perfectly match any particular known or inferred history, but instead
demonstrate the effect of reciprocal introgression following population
divergence -- a population of size $N_e=10\,000$ splits into two, on remaining
size $10\,000$ and the other shrinking to size $1\,000$. They remain isolated
for $2N_e$ generations (or 500 thousand years, assuming an average generation
time of 25 years) and then introgression occurs from one branch to the other,
contributing $5\%$ ancestry to the recipient population
(Figure~\ref{toy-admixture}A,B).

@@

In each scenario, we track genetic diversity and phenotypic variance before and
after admixture to understand allele frequency and haplotype dynamics. The
trait optimum is $0$ in all branches, and the strength of selection $V_S=1$
remains constant. \aprcomment{Justified by equivalence under rescaling
\citep{simons2018population}.}

@@

We expect genetic variance to increase after introgression. The amount that
genetic variance increases depends on the allelic differences accumulated
between populations and the effects of those alleles. Assuming no linkage,
dominance or epistasis, \(V_G=\sum_l 2p_l(1-p_l)a_l^2\). After admixture, with
proportion $f$ contributed by the source population (labeled 0) into the focal
population (labeled 1), \(p_l = fp_{l,0} + (1-f)p_{l,1}\). Plugging into the
expression for $V_G$, and after some simple algebra (Appendix~AX), we can
express the expected genetic variance directly after admixture as \[V_G =
fV_{G,0} + (1-f) V_{G,1} + 2f(1-f)\sum_l F_{2,l} a_l^2,\] where $F_2 = (p_0 -
p_1)^2$ is the squared difference in allele frequencies at a locus
\citep[e.g.,][]{peter2016admixture}. This result is known
\citep[e.g.,][]{tufto2000quantitative}, showing that additive variance is equal
to that in the source populations weighted by their contributions, plus a term
that depends on the divergence at trait-affecting loci between the populations
weighted by the quadratic factor $2f(1-f)$.

@@

\begin{figure}[t!]
    \centering
    \includegraphics{../figures/model_comparison.SD_0.01.mu_0.001_0.025}
    \caption{}
    \label{fig:linkage-deviations}
\end{figure}

@@

\begin{figure}[t!]
    \centering
    \includegraphics{../figures/introgression_deserts.SD_0.02.pdf}
    \caption{
        \textbf{}
    }
    \label{fig:deserts}
\end{figure}

@@

\begin{itemize}
    \item Stabilizing selection with a constant optimum is an appropriate
        null model for the dynamics of trait variation and selection on
        alleles underlying those traits
    \item Stabilizing selection results in population genetic predictions
        for patterns of introgressed variation in and around functional
        regions that differ from both load- and incompatibility-based arguments
        \begin{itemize}
            \item Load-based models predict that haplotypes from the less ``loaded''
                population will be favored under introgression in either direction
                -- thus, regions of depleted ancestry after introgression in one
                direction will not be expected to also be introgressed ancestry
                deserts after admixture in the opposite direction
            \item \citet{mueller1942} originally pointed out that hybrid
                incompatibilities should not form at single loci -- classical
                incompatibilities require two or more loci, which are not
                expected to be tightly linked to one another. Thus, under a
                model of hybrid incompatibilities, the minor parent ancestry
                is again selected against, but those loci will be in different
                regions. Again, deserts are therefore not expected to overlap
                if selection is largely due to incompatibilities
            \item Instead, stabilizing selection causes introgressed DNA to be
                depleted at the same loci under bi-directional gene flow
                (could be thought of as single-locus incompatibility effects --
                but these are not incompatibilities in the classical sense,
                as the same effects occur even without admixture)
            \item A prediction we could make is that, if current models are correct
                that there has been introgression in both directions between
                \emph{H. sapiens} and Neanderthals, introgressed ancestry deserts
                should overlap with each other more often than expected by chance
            \item See Figure~\ref{fig:ancestry-deserts}
        \end{itemize}
    \item Shared deserts are most likely to form where moderate- or large-effect
        alleles have fixed in one population or the other -- this could provide
        an approach to map\dots?
    \item Selection in structured populations with ongoing migration or
        meta-populations -- SFS approach can handle this -- prospects for the
        future
    \item Some conjectures about hybrid zones
    \item Pleiotropy\dots!!!
\end{itemize}
